---
output:
  pdf_document: default
  html_document: default
---
# Making Red Wine--Art of Science?
## by Y. Jeremy Dai

========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
setwd("/Users/jeremydai/Dropbox/Udacity_Project/EDA Red Wine") 
library(ggplot2)
library(gridExtra)
library(GGally)
library(dplyr)
library(memisc)
```

```{r echo=FALSE,warning=FALSE, Load_the_Data}
wine<- read.csv("wineQualityReds.csv")
```

> **Intro**: This report explores a dataset containing qualities and attributes
for 1,599 red wines. The goal is to find the how the wine quality is related to
these objective attributes.

# Univariate Plots Section

```{r echo=FALSE,warning=FALSE, summaryofattributes}
dim(wine)
summary(wine[,2:13])
print('Quality frequency')
table(wine$quality)
```

Our dataset consists of 13 variables, with 1,599 observations.

The first varible is the number of the observation. Varible 2-12 are the 
objective attributes. The last varible is the output (quality).

```{r echo=FALSE,warning=FALSE, bargraph}
ggplot(aes(x= as.factor(quality)),data=wine)+geom_bar(fill ='dark red')
```

We find that there are much more normal wines (of average quality) than 
excellent or poor ones (of high or low quality). One of the reasons may be that
the quality is the median of at least 3 evaluations made by wine experts. 
Therefore, the distribtuion gets narrower after the measurements being averaged.

First we are going to look into the attributes regarding acid:

   1 - fixed acidity (tartaric acid - g / dm^3) 
   
   2 - volatile acidity (acetic acid - g / dm^3) 
   
   3 - citric acid (g / dm^3) 

```{r echo=FALSE,warning=FALSE, acid}
# Use dark red as the theme color to match the red wine appearance
plot.hist <- function(x.var, df, bw){
  ggplot(df, aes_string(x = x.var)) +
    geom_histogram(binwidth = bw,fill ='dark red')
}

p1<-plot.hist('fixed.acidity', wine, 1) + xlab('fixed.acidity')
p2<-plot.hist('volatile.acidity', wine, 0.02) + xlab('volatile.acidity')
p3<-plot.hist('citric.acid', wine, 0.02) + xlab('citric.acid')

grid.arrange(p1,p2,p3,ncol=1)
print('citric.acid=0?')
summary(wine$citric.acid==0)
```

Both fixed acidity and volatile acidity have bell-shaped-like distribution. 
Citric acid distribution, on the other side, is right skewed and looks like a
bi-mode distrbution. I wonder why there are more than one peak.

Also, there are wines without any citric acid (132 out of 1599).

Next, we are going to look into the attributes regarding sweetness:

   4 - residual sugar (g / dm^3) 
   
   11 - alcohol (% by volume) 


```{r echo=FALSE, warning=FALSE,sweet}
p1<-plot.hist('residual.sugar', wine, 0.5)

plot.box <- function(x.var, df){
  ggplot(df, aes_string(y = x.var, x = '1')) +
  geom_boxplot(color = 'dark red') +
  geom_jitter(aes_string(x = '1', y = x.var), alpha = 0.04) +
  labs(y = x.var , x = '')+
  coord_flip()

}

p2 <- plot.box('residual.sugar', wine)

grid.arrange(p1,p2)

plot.hist('alcohol', wine, 0.2) + xlab('alcohol')

```


We can see that both residual sugar and alcohol distributions are skewed to the
right. Residual sugar distribution has a long tail. However, wines do not differ
a lot in alcohol concentration but there are some outliers in residual sugar 
distributions.Some wines are much sweeter than the average. 

After transforming x axis, we find that most wine have residual sugar of around 
2 g /L. Most of wines in this study are dry (0-9 g/L RS) and few wines are
off-dry (9-18 g/L RS). There is no sweet wine(>18 g/L RS) in this study.

Next, salinity is explored.

  5 - chlorides (sodium chloride - g / dm^3 

```{r echo=FALSE, warning=FALSE,salt}
p1 <- plot.hist('chlorides', wine, 0.02)

p2 <- plot.box('chlorides', wine)

grid.arrange(p1,p2)
```

Similar to residual sugar, chlorides distribution is highly right skewed with 
some outliers to the right. Some wines are much saltier than the average.

Next, we will look into the attributes regarding bitterness.

   6 - free sulfur dioxide (mg / dm^3) 
   
   7 - total sulfur dioxide (mg / dm^3) 
   
   10 - sulphates (potassium sulphate - g / dm3) 

```{r echo=FALSE, warning=FALSE,bitter}
p1 <- plot.hist('free.sulfur.dioxide', wine, 5) 
p2 <- plot.box('free.sulfur.dioxide', wine)
grid.arrange(p1,p2)

p1 <- plot.hist('total.sulfur.dioxide', wine, 10) 
p2 <- plot.box('total.sulfur.dioxide', wine)
grid.arrange(p1,p2)

p1 <- plot.hist('sulphates', wine, 0.02) 
p2 <- plot.box('sulphates', wine)
grid.arrange(p1,p2)
```

All of them are right skewed. There are some outliers to the right.

Finally, we will look into the density and PH.

   8 - density (g / cm^3)
   
   9 - pH


```{r echo=FALSE, warning=FALSE,other}
plot.hist('density', wine, 0.001) + xlab('density')
plot.hist('pH', wine, 0.05) + xlab('pH')
```

As shown in the figures, both the pH and density form a bell-shape distribution.
The density of most wine is smaller than 1.000. This is not surprising since 
alcohol is lighter than water. Also, pH of most wines is within 3.0 to 3.5. 
This is due to the exisitence of fixed, volatile and citric acid. 

# Univariate Analysis

### What is the structure of your dataset?
There are 12 features for 1,599 wines in our data. Other than quality, the
features are measured as a continuous variable. The classes of quality is 
ordered : between 0 (very bad) and 10 (very excellent). 

Notably, the wine quality rating is not proportional. A wine of quality 6 is not
twice better than a wine of quality 3. Also, the wine sample in this study is 
highly imbalanced with most wines of quality equal to 5 or 6.

### What is/are the main feature(s) of interest in your dataset?
The main feature of interest is the quality fo the wine.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
The chemicals contributing to the acid, sweet, bitter or salty taste are
the features that will be explored in this study.

### Did you create any new variables from existing variables in the dataset?
Not yet.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
There are some outliers in most distributions. I do not want to remove them 
since they may contribute to the few instances of poor/good quality wines.


# Bivariate Plots Section

```{r echo=FALSE,warning=FALSE, cor}
cor(wine[,2:13])
```


```{r echo=FALSE, warning=FALSE,summary}
ggpairs(data=wine[,2:13], lower = list(continuous = wrap("points", alpha = 0.1,
                                         shape = I('.'),color='dark red')),
                               upper = list(continuous = wrap("cor", size = 3)))

```

Based on the correlation matrix and point plots, there is no strong correlation 
(cor>0.7) between any of two parameters. 

The correlation is low for the quality with pH, residual sugar or free 
sulfurdioxide. The attribute with strongest correlation with the quality is 
alcohol.

```{r echo=FALSE, warning=FALSE,ph_fixacid}
ggplot(wine,aes(pH,fixed.acidity)) + geom_point(alpha=0.1,color='dark red')
ggplot(wine,aes(density,fixed.acidity)) + 
  geom_point(alpha=0.1,color='dark red')
```

pH tends to be higher with lower fixed acidity. Density tends to be higher with
higher fixed acidity.

A new varible to categorize wine quality is created to balance the sample a 
little bit. 

1. high quality: quality >=7

2. medium high quality: quality equal to 6

3. medium low quality: quality equal to 5 

4. low quality: quality <=4

Since there are way more medium quality wines, the heights of histograms for 
wines of different quality are quite different. Therefore, y axis for histgram 
is transformed using 'sqrt' so we can see the distribution of low/high quality
wine.

```{r echo=FALSE, warning=FALSE,density}
wine$quality_f<-'Medium Low'
wine$quality_f<-with(wine, ifelse(quality==6,'Medium High', quality_f))
wine$quality_f<-with(wine, ifelse(quality<=4,'Low', quality_f))
wine$quality_f<-with(wine, ifelse(quality>=7,'High',quality_f)) 

wine$quality_f <- factor(wine$quality_f, 
                         levels = c("High", "Medium High", "Medium Low", "Low"))

for(i in 1:11){
  p <-ggplot(wine, aes_string(x=names(wine)[i+1],color='quality_f', 
                                   group='quality_f')) + 
  geom_density()+
  scale_color_brewer(type='seq') +
  labs(colour = "Quality")
  print(p)
}
```



```{r echo=FALSE, warning=FALSE,frequency}
plot.freq <- function(x.var, df){
  ggplot(df, aes_string(x=x.var))+
              geom_histogram(fill ='dark red',position="identity")+
              coord_trans(y = "sqrt")+
              facet_wrap(~quality_f)
}

plot.freq('sulphates',wine)
plot.freq('citric.acid',wine)
plot.freq('volatile.acidity',wine)
plot.freq('alcohol',wine)
```

From the density curves and histograms, we can see that:

It is strange that there are some a few high quality wines containing a lot of
total sulfur dioxide (>250 mg / dm^3), which gives a bitter taste, while most
of wines have low total sulfur dioxide (<160 mg / dm^3)

Also, better wines tend to have more sulphates. However, when the sulphates is
more than 1.5 g / dm3, there is a big chance that the wine is not of high 
quality.

When the citric acid is low, there is higher chance that the wine is of 
low quality. When the citric acid is around 0.4 g / dm^3, the wine is more
likely to be high quality.

Wine with high quality tends to have high alcohol percentage.

Most wines that have high chlorides (>0.2 g / dm^3) are low/medium quality.

The average volatile acidity percentage is lower for high quality wines, higher
for low quality wines.

```{r echo=FALSE,warning=FALSE, mean}
quality_groups <- group_by(wine, quality)
wine_by_quality <- summarise(quality_groups,
             sugar_mean = mean(residual.sugar),
             sulphate_mean = mean(sulphates),
             alcohol_mean=mean(alcohol),
                  n = n())

p1<-ggplot(wine_by_quality, aes(quality, sugar_mean)) + geom_line(color='dark red')
p2<-ggplot(wine_by_quality, aes(quality, sulphate_mean)) + geom_line(color='dark red')
p3<-ggplot(wine_by_quality, aes(quality, alcohol_mean)) + geom_line(color='dark red')
grid.arrange(p1,p2,p3,ncol=1)

```

```{r echo=FALSE, warning=FALSE,Boxplots}
p1<-ggplot(wine,aes(as.factor(quality),alcohol)) + geom_boxplot()
p2<-ggplot(wine,aes(as.factor(quality), residual.sugar)) + geom_boxplot()+
   coord_cartesian(ylim=c(0,4))
grid.arrange(p1,p2,ncol=1)
```

From the boxplots and line diagrams for the average attributes of wine of 
different qualities, we find that:

1. Wine of medium quality has the lowest residual sugar. However, the difference
is not large.

2. The average sulphate increases when the wine quality improves.

3. The quality tends to improve when the alcohol concentrate is higher.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

If sulphates is more than 1.5 g / dm3 or citric acid is less than 0.25 g / dm3,
the wine is more likely to be of low quality.

When the citric acid is around 0.4 g / dm^3 or the alcohol percentage is higher
than 11, the wine is more likely to be high quality.

Most wines that have high chlorides (>0.2 g / dm^3) are low/medium quality.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
Density tends to be higher with higher fixed acidity. After googling, we find
that tartaric acid is denser than water and alcohol.

### What was the strongest relationship you found?
The strongest relationship is between fixed.acidity and pH. This is not 
surprising since pH measure how much acid the liqucid contains.

# Multivariate Plots Section

```{r echo=FALSE, warning=FALSE,sugar_fixacid}
ggplot(wine,aes(x=residual.sugar,y=fixed.acidity)) + 
  geom_point(aes(color=quality_f)) + 
  scale_color_brewer(type="seq")

ggplot(subset(wine,residual.sugar<8),aes(x=residual.sugar,y=fixed.acidity)) + 
  geom_point(aes(color=quality_f)) + 
  scale_color_brewer(type="seq")
```

There are some obvious outlier in residual sugar distribution, most of them are
of medium quality. Also, if we exlcude these outliers, there is more variation 
in fixed acidity axis than reisudal sugar axis.

```{r echo=FALSE, warning=FALSE,alcohol_citricacid}
ggplot(wine,aes(x=alcohol,y=citric.acid)) + 
  geom_point(aes(color=quality_f)) + 
  scale_color_brewer(type="seq")

ggplot(wine,aes(x=alcohol,y=sulphates)) + 
  ylim(c(0.2,1.2))+
  geom_point(aes(color=quality_f)) + 
  scale_color_brewer(type="seq")
```

When the alcohol and citric acid/sulphates are both high, there is high change that the 
wine is of high quality.

```{r echo=FALSE, warning=FALSE,Linear_Model}
m1 <- lm(I(quality) ~ I(alcohol), data = wine)
m2 <- update(m1, ~ . + citric.acid)
m3 <- update(m2, ~ . + volatile.acidity)
m4 <- update(m3, ~ . + sulphates)
mtable(m1, m2, m3, m4, sdigits = 3)
```

To predict the wine quality, four attributes are included in the linear model:
alcohol, citric.acid , volatile.acidity and sulphates. Even after we include
all of them, the adj. R-square is only 0.334. This is not a good result.

```{r echo=FALSE, warning=FALSE,pca}
wine.pca <- prcomp(wine[2:12], center = TRUE, scale. = TRUE) 
print(wine.pca)
plot(wine.pca, type = "l")
summary(wine.pca)
```

Based on PCA results, the top four principle components can account for 70
% of the variance (0.2817+ 0.1751+ 0.1410+ 0.1103 = 0.7081).

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?
When the alcohol and citric acid are both high, there is high change that the 
wine is of high quality.

### Were there any interesting or surprising interactions between features?
If we remove the outliers in residual sugar, there is not much variation. No
wonder residual sugar does not have a big influence on wine quality.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.
Yes.  

A linear model is created with alcohol, citric.acid , volatile.acidity and
sulphates as predictors. The result is not good for the following reasons:

1. We consider quality as a continuous attribute in the model while it should be
a categorical variable.

2. A linear model may not be the best fit.

3. The sample is highly unbalanced.

Also, we conduct principle component analysis. The top four principle components
can only account for 70 % of the variance

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, warning=FALSE,Plot_One}
ggplot(aes(x=as.factor(quality)),data=wine)+geom_bar(fill ='dark red')+
  ggtitle('Distribution of Wine Quality') +
  xlab('Wine Quality Rating')
  theme(plot.title = element_text(hjust = 0.5))
```

### Description One
The wine sample is highly unbalanced with few low/high quality wines and a 
majority of medium quality wines (5 or 6). This is not surprising. Either 
normal distirbutions or our experiences indicates that, for many product, 
most of them are medium while a few of them are either awesome or terrible. For
example, there are a few best cell phone models like iPhone X and Sumsung 
Note 8. There are a lot of 'okay' phone models from Asian countries. Also, a few
non-smart phones still exist.

### Plot Two
```{r echo=FALSE, warning=FALSE,Plot_Two}
ggplot(wine, aes(x=sulphates,color=quality_f, group=quality_f))+
  xlab('potassium sulphate - g / dm3 ')+ 
  ylab(' kernel density estimate' ) + 
  geom_density()+
  scale_color_brewer(type='seq') +
  ggtitle('Kernel Density Distribution for Potassium Sulphate') +
  theme(plot.title = element_text(hjust = 0.5)) 

```

### Description Two
Wine with high quality(7,8) has a higher medium potassium sulphate concentration.
However, if there are too much potassium sulphate(too bitter), it is very likely
that the wine is of low quality (3,4).

### Plot Three
```{r echo=FALSE, warning=FALSE,Plot_Three}
ggplot(wine,aes(x=alcohol,y=citric.acid)) +
  xlab('alcohol (% by volume)')+
  ylab('citric acid (g / dm^3) ') + 
  geom_point(aes(color=quality_f)) + 
  scale_color_brewer(type="div") + 
  ggtitle('Wine Quality by Alcohol and Citric Acid') +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(color="Wine Quality")

```

### Description Three
There are not many high quality wines in the sample. However, if the wine is
high in both alcohol and citirc acid, there is a good chance that the wine of
high quality (7 or 8).
------

# Reflection
This study explores the physical attributes of red wines and how they affect
the wine quality.  This sample is highly imbalanced, which means there are
fewer wines with high/low quality.

In order to better understand the relationship between wine quality and 
objective attributes, we used the following methods:

1. Use density diagram to ignore the size in each quality group

2. Group the wine of quality of 3 and 4; group the wine of quality of 7 and 8

3. Transform y axis

4. Use average measurement of each quality group


In this study, we find that the wine tends to be of high 
quality if it is high in alcohol,sulphates and citric acid. However, if these
substances are extremely high, the wine is likely to be of low quality.

Also, we find that the linear regression is not a good model to find the 
underneath pattern. PCA shows that even top four components combined only 
account for around 70% of the variance. It looks like it is hard to make a good
wine if you just adjust objective attributes. Making wines is more like an art
instead of a simple techonology that you only need to play with numbers. 

The next step would be tring non-linear models. We can also use machine learning
- the most sophisticated and currently most popular approach. 

> **Citation**: This dataset is public available for research. The details are 
described in [Cortez et al., 2009]. Reference:
  P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
  Modeling wine preferences by data mining from physicochemical properties.
  In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.

> **Available at: **
  [@Elsevier] http://dx.doi.org/10.1016/j.dss.2009.05.016
  [Pre-press (pdf)] http://www3.dsi.uminho.pt/pcortez/winequality09.pdf
  [bib] http://www3.dsi.uminho.pt/pcortez/dss09.bib
